---
title: "Import dataset of DMTA on Devionan sharks"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: Sharks_1_Import.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---


```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
This script formats the output of the resulting files from applying surface texture analysis to a sample of Devonian shark teeth.
The script will:

1. Read in the original files  
2. Format the data  
3. Write ODS file and save R objects ready for further analysis in R  

```{r}
dir_in  <- "analysis/raw_data"
dir_out <- "analysis/derived_data"
```

Raw data must be located in "`r paste0("./", dir_in)`".  
Formatted data will be saved in "`r paste0("./", dir_out)`".

The knit directory for this script is the project directory.


---


# Load packages
```{r}
pack_to_load <- sort(c("R.utils", "readODS", "tidyverse", "rmarkdown", "knitr", "grateful"))
sapply(pack_to_load, library, character.only = TRUE, logical.return = TRUE) 
```


---


# Get names and path all files 
```{r}
info_in <- list.files(dir_in, pattern = "\\.csv$", full.names = TRUE)
info_in
```


---


# Read and format data
## Read in CSV files
```{r}
# Loop through list of CSV files and read in
sharks <- info_in %>% 
  
           # Subset relevant CSV files
           .[grepl("x.csv", .)] %>% 
  
           # Read in all relevant CSV files
           lapply(function(x) read.csv(x, header = FALSE, na.strings = "*****", 
                                       fileEncoding = 'WINDOWS-1252')) %>%
  
           # rbind together
           do.call(rbind, .)
```


## Select relevant columns and rows
```{r}
sharks_keep_col  <- c(4, 26:55, 57:58, 61:64)                 # Define columns to keep
sharks_keep_rows <- which(sharks[[1]] != "#")                 # Define rows to keep
sharks_keep      <- sharks[sharks_keep_rows, sharks_keep_col] # Subset rows and columns
```


## Identify results using frame numbers
```{r}
frames <- as.numeric(unlist(sharks[1, sharks_keep_col]))
ID <- which(frames %in% c(2, 11))
ISO <- which(frames == 15)
furrow <- which(frames == 16)
diriso <- which(frames %in% 17:18)
SSFA <- which(frames %in% 19:20)
```


## Add headers
```{r}
# Get headers from 2nd row
colnames(sharks_keep) <- sharks[2, sharks_keep_col] %>% 
  
                         # Convert to valid names
                         make.names() %>% 
  
                         # Delete repeated periods
                         gsub("\\.+", "\\.", x = .) %>% 
  
                         # Delete periods at the end of the names
                         gsub("\\.$", "", x = .)
  
#
colnames(sharks_keep)[ISO] <- strsplit(names(sharks_keep)[ISO], ".", fixed = TRUE) %>% 
                              sapply(`[[`, 1)
#
colnames(sharks_keep)[SSFA] <- gsub("^([A-Za-z0-9]+\\.)+", "", colnames(sharks_keep)[SSFA])

# Edit headers for name of surfaces and non-measured point ratios
colnames(sharks_keep)[ID] <- c("Name", "NMP")
```


## Extract units
```{r}
# Filter out rows which contains units
n_units <- filter(sharks, V4 == "<no unit>") %>% 
          
           # Keep only unique/distinct rows of units
           distinct() %>% 
  
           # Number of unique/distinct rows of units
           nrow()

if (n_units != 1) {
  stop("The different datasets have different units")
} else {
  # Extract unit line from 3rd row for considered columns
  sharks_units <- unlist(sharks[3, sharks_keep_col[-1]])

  # Get names associated to the units
  names(sharks_units) <- colnames(sharks_keep)[-1]

  # Combine into a data.frame for export
  units_table <- data.frame(variable = names(sharks_units), units = sharks_units)
}
```


## Split column 'Name' 
```{r}
sharks_keep[c("Specimen", "Tooth", "Location", "Objective", "Measurement")] <- str_split_fixed(
                                                                               sharks_keep$Name, "_", n = 5)
```


## Convert to numeric
```{r}
# Replace "-nan(ind)" with NA in column HAsfc81
sharks_keep$HAsfc81 <- na_if(sharks_keep$HAsfc81, "-nan(ind)")

# Convert all parameter variables to numeric
sharks_keep <- type_convert(sharks_keep)
```


## Add column for NMP categories
Here we define 3 ranges of non-measured points (NMP):

- $<$ 10% NMP: "<10%"  
- $\ge$ 10% and $<$ 20% NMP: "10-20%"  
- $\ge$ 20% NMP: "≥20%"
 

```{r}
# Create new column and fill it
sharks_keep[sharks_keep$NMP < 10                        , "NMP_cat"] <- "<10%"
sharks_keep[sharks_keep$NMP >= 10 & sharks_keep$NMP < 20, "NMP_cat"] <- "10-20%"
sharks_keep[sharks_keep$NMP >= 20                       , "NMP_cat"] <- "≥20%"

# Convert to ordered factor
sharks_keep[["NMP_cat"]] <- factor(sharks_keep[["NMP_cat"]], levels = c("<10%", "10-20%", "≥20%"), ordered = TRUE)
```

The number of height maps for each objective in each NMP category is:  
```{r}
table(sharks_keep[c("NMP_cat", "Objective")])
```


## Read in CSV files of measurement positions
```{r}
meas_pos <- info_in %>% 
            .[grepl("positions", .)] %>% 
            lapply(function(x) read.csv2(x, fileEncoding = 'WINDOWS-1252')) %>%
            do.call(rbind, .)
str(meas_pos)
head(meas_pos)
```


## Merge measurement positions with dataset
```{r}
sharks_keep_pos <- merge(sharks_keep, meas_pos, by = "Name", all.x = TRUE)
```


## Re-order columns and add units as comment 
```{r}
sharks_final <- select(sharks_keep_pos, Specimen:Measurement, Position, NMP, NMP_cat, Sq:HAsfc81)
comment(sharks_final) <- sharks_units
```

Type `comment(sharks_final)` to check the units of the parameters.


## Check the result
```{r}
str(sharks_final)
head(sharks_final)
```


---


# Save data
## Create file names
```{r}
sharks_ods <- paste0(dir_out, "/DMTA-Ctenacanths.ods")
sharks_rbin <- paste0(dir_out, "/DMTA-Ctenacanths.Rbin")
```


## Write to ODS and Rbin
```{r}
write_ods(list("data" = sharks_final, "units" = units_table), path = sharks_ods) 
saveObject(sharks_final, file = sharks_rbin) 
```

Rbin files (e.g. `DMTA-Ctenacanths.Rbin`) can be easily read into an R object (e.g. `rbin_data`) using the following code:
```{r, eval = FALSE}
library(R.utils)
rbin_data <- loadObject("DMTA-Ctenacanths.Rbin")
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = c(pack_to_load, "base"), output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "Sharks_1_Import")
knitr::kable(pkgs_cite)
```


## References

