---
title: "Summary statistics on dataset of DMTA on Devionan sharks"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: Sharks_2_Summary-stats.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---


```{r Knitr Options, include = FALSE}
	knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
This script computes standard descriptive statistics for each group.  
The groups are based on:

* Specimen  
* Tooth position  
* NMP_cat

It computes the following statistics:  

* sample size (n = `length`)  
* smallest value (`min`)  
* largest value (`max`)
* mean  
* median  
* standard deviation (`sd`)

```{r}
dir_in  <- "analysis/derived_data"
dir_out <- "analysis/stats"
```

Input Rbin data file must be located in "`r paste0("~/", dir_in)`".  
Summary statistics table will be saved in "`r paste0("~/", dir_out)`".

The knit directory for this script is the project directory.


---


# Load packages
```{r}
pack_to_load <- c("doBy", "grateful", "knitr", "openxlsx", "R.utils", "rmarkdown", "tidyverse")
sapply(pack_to_load, library, character.only = TRUE, logical.return = TRUE) 
```


---


# Read in data
## Get name and path of input file 
```{r}
info_in <- list.files(dir_in, pattern = "\\.Rbin$", full.names = TRUE)
info_in
```


## Read in Rbin file
```{r}
sharks <- loadObject(info_in)
str(sharks)
```


---


# Summary statistics
## Create function to compute the statistics at once
```{r}
nminmaxmeanmedsd <- function(x){
	y <- x[!is.na(x)]     # Exclude NAs
	n_test <- length(y)   # Sample size (n)
	min_test <- min(y)    # Minimum
	max_test <- max(y)    # Maximum
	mean_test <- mean(y)  # Mean
 	med_test <- median(y) # Median
 	sd_test <- sd(y)      # Standard deviation
 	out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test) # Concatenate
 	names(out) <- c("n", "min", "max", "mean", "median", "sd")         # Name values
 	return(out)                                                        # Object to return
}
```


## Define grouping and numerical variables to use
```{r}
# Define grouping variables: Specimen + NMP_cat
grp1 <- c("Specimen", "NMP_cat")

# Define grouping variables: Specimen + Tooth
grp2 <- c("Specimen", "Tooth")

# Define grouping variables: Specimen + Tooth + NMP_cat
grp3 <- c("Specimen", "Tooth", "NMP_cat")
```

The following grouping variables will be used:  

* First set `grp1`
```{r, echo = FALSE}
cat(colnames(select(sharks, all_of(grp1))), sep = "\n")
```
* Second set `grp2`
```{r, echo = FALSE}
cat(colnames(select(sharks, all_of(grp2))), sep = "\n")
```
* Third set `grp3`
```{r, echo = FALSE}
cat(colnames(select(sharks, all_of(grp3))), sep = "\n")
```

All numerical variables except `NMP` will be used: 
```{r, echo = FALSE}
cat(colnames(select(sharks, where(is.numeric))[-1]), sep = "\n")
```


## Compute summary statistics
```{r}
# grp1
# Create formula based on grouping variables in grp1
stats_grp1 <- as.formula(paste(".~", paste(grp1, collapse = "+"))) %>% 
  
              # calculate group-wise stats
              summaryBy(data = sharks, FUN = nminmaxmeanmedsd)

# Same with grp2
stats_grp2 <- as.formula(paste(".~", paste(grp2, collapse = "+"))) %>% 
              summaryBy(data = sharks, FUN = nminmaxmeanmedsd)

# Same with grp3
stats_grp3 <- as.formula(paste(".~", paste(grp3, collapse = "+"))) %>% 
              summaryBy(data = sharks, FUN = nminmaxmeanmedsd)
```


## Add units
```{r}
# Extract units from comment(sharks) and exclude first row (NMP)
units_stats <- data.frame(variable = names(comment(sharks)), units = comment(sharks))[-1,]
```


---


# Write results to XLSX
```{r}
write.xlsx(list(Specimen_NMP = stats_grp1, Specimen_Tooth = stats_grp2, Specimen_Tooth_NMP = stats_grp3, 
                units = units_stats), file = paste0(dir_out, "/DMTA-Ctenacanths_summary-stats.xlsx"))
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
cite_packages(pkgs = c(pack_to_load, "base"), output = "paragraph", 
              include.RStudio = TRUE, out.dir = "analysis/scripts", bib.file = "Sharks_2_Summary-stats")
```


## References

