---
title: "Plots of SSFA variables for the dataset of DMTA on Devionan sharks"
author: "Ivan Calandra"
date: "`r Sys.time()`"
output:
  github_document: 
    toc: true
    toc_depth: 2
    html_preview: false
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
The script plots all SSFA variables for the Devonian shark dataset.   

```{r}
dir_in  <- "derived_data"
dir_out <- "plots"
```

Input Rbin data file must be located in "`r paste0("~/", dir_in)`".  
Plots will be saved in "`r paste0("~/", dir_out)`".

The knit directory for this script is the project directory.


---


# Load packages
```{r}
pack_to_load <- c("R.utils", "ggplot2", "RColorBrewer", "tidyverse")
sapply(pack_to_load, library, character.only = TRUE, logical.return = TRUE)
```


---


# Read in data
## Get name and path of input file 
```{r}
info_in <- list.files(dir_in, pattern = "\\.Rbin$", full.names = TRUE)
info_in
```


## Read in Rbin file
```{r}
all_data <- loadObject(info_in)
str(all_data)
```


---


# Define variables
Here we define which columns are used for the plots.  

```{r}
# Combine columns species and specimen into one to be used to group on the x-axis
all_data[, "Species.Specimen"] <- paste(all_data[["Species"]], all_data[["Specimen"]], sep = ".")
x_var <- "Species.Specimen"

# Surface parameters to plot
y_var <- c("Asfc", "epLsar", "Smfc", "HAsfc9")

# colors
grp_colors <- "NMP_cat"

# shapes
grp_shapes <- "Location"
```

The following variables will be used:  
```{r}
x_var
y_var
grp_colors
grp_shapes
```


---


# Exclude surfaces with NMP ≥ 20%
Surfaces with more than 20% NMP should not be analyzed.

```{r}
data_nmp0_20 <- filter(all_data, NMP_cat != "≥20%")
data_nmp0_20$NMP_cat <- factor(data_nmp0_20$NMP_cat)
str(data_nmp0_20)
```


---


# Plot each surface parameter in a boxplot
## Pivot to longer format for facet plots
```{r}
data_long <- select(data_nmp0_20, all_of(c(x_var, y_var, grp_colors, grp_shapes))) %>%
             pivot_longer(all_of(y_var), names_to = "parameter", values_to = "value")
```


## Plot
```{r}
# set up plot
p_box <- ggplot(data_long, aes_string(x = x_var, y = "value")) +

         # Boxplots:
         # hide outliers (all points are shown with geom_point() below) 
         geom_boxplot(outlier.shape = NA) +
  
         # Points:
         # Add layers of shapes and colors for points 
         # Jitter points
         geom_point(mapping = aes_string(color = grp_colors, shape = grp_shapes), position = "jitter", size = 2) +
  
         # Remove y-axis label
         labs(y = NULL) + 
  
         # Choose a light theme
         theme_classic() +
  
         # Wrap around parameters with free y-scales
         facet_wrap(~ parameter, scales = "free_y") +

         # The qualitative 'Set2' palette of RColorBrewer is colorblind friendly and has 8 colors
         scale_color_brewer(palette = 'Set2')

# Print plot
print(p_box)
```


## Save plot
```{r}
ggsave(plot = p_box, filename = "DMTAsharks_EAVP_100x_SSFAboxplots.pdf", 
       path = dir_out, width = 240, height = 190, units = "mm")
```


---


# Plot anisotropy vs. complexity
## Plot
```{r}
# set up plot
p_bi <- ggplot(data_nmp0_20, aes(x = Asfc, y = epLsar)) +
        
        # Scatterplot
        geom_point(mapping = aes_string(color = x_var), size = 4) +
  
        # Adjust axes labels
        labs(x = "Complexity (Asfc)", y = "Anisotropy (epLsar)") +
  
        # The qualitative 'Set1' palette of RColorBrewer is colorblind friendly and has 9 colors
        scale_color_brewer(palette = 'Set1') +
  
        # Choose a light theme
        theme_classic()

# Print plot
print(p_bi)
```


## Save plot
```{r}
ggsave(plot = p_bi, filename = "DMTAsharks_EAVP_100x_epLsar-Asfc.pdf", 
       path = dir_out, width = 240, height = 190, units = "mm")
```

---


# sessionInfo() and RStudio version

```{r}
sessionInfo()
```

RStudio version `r readLines("scripts/Sharks_0_RStudioVersion.txt", n = 1)`.


---


# Cite R packages used
Does not work yet


---

END OF SCRIPT
